Human capital mediates natural selection in contemporary humans

### 2021-12-14 why-natural-selection.Rmd
Add analysis of natural selection's effect on true PSEA's correlation with income; tweak intro; add function to do mediation analysis using 1973 ROSLA as an exogenous shift to education

library(drake)
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)

@@

This study uses data from UK Biobank [@bycroft2018uk] to learn more about
contemporary natural selection. We test for natural selection on 33 different
polygenic scores by estimating their correlation with fertility. We extend the
analysis over two generations, using data on respondents' number of siblings as
well as their number of children. This is interesting because consistent natural
selection over multiple generations could lead to substantive effects in the
long run. We then examine reproductive rates in different subgroups of the
population. We find selection effects on many polygenic scores. Effects are
largely consistent across generations.

Next, we examine correlations with fertility in different subgroups. Across
the board, selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with more lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed.

@@

@@

cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100
```

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
proportionate increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. Since
many polygenic scores are predictive of disease risk, they could also
potentially increase health inequalities. In general, the evolutionary history
of anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

@@

# Proofs

>>>>

library(drake)
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)
loadNamespace("scales")

@@

This study uses data from UK Biobank [@bycroft2018uk] to learn more about
contemporary natural selection. We test for natural selection on 33 different
polygenic scores by estimating their correlation with fertility. We extend the
analysis over two generations, using data on respondents' number of siblings as
well as their number of children. This is interesting because consistent natural
selection over multiple generations could lead to substantive effects in the
long run. Next, we examine correlations with fertility in different subgroups. Across
the board, selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with more lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed.

@@

Lastly, we discuss the effects of natural selection. While measured effects on
mean genetic scores are small, natural selection may substantially increase the correlation
between genetic scores and income, i.e. the "genetic lottery" [@harden2021genetic].

@@

cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100

drake::loadd(res_ineq_ea3)

r2_income_true_psea <- scales::percent(res_ineq_ea3$r2_income_true_psea, 
                                         accuracy = 0.1)
r2_income_true_psea_wt <- scales::percent(res_ineq_ea3$r2_income_true_psea_wt, 
                                            accuracy = 0.1)
```

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
proportionate increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)).

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
income attributable to the "true" polygenic score for educational attainment as
`r r2_income_true_psea` among respondents. Counterfactually holding environmental
conditions constant, natural selection would raise this proportion to 
`r r2_income_true_psea_wt` among respondents' children. Similarly, since many
polygenic scores are predictive of disease risk, they could potentially increase
health inequalities. In general, the evolutionary history of anatomically modern
humans is related to disease risk [@benton2021influence]; understanding the role
of contemporary natural selection may aid in mapping the genetic architecture of
current health disparities.

@@

## Estimating the effect of natural selection on the contribution of PSEA to inequality

```{r calc-load-ineq-ea3}

res <- drake::readd(res_ineq_ea3)

prop_effect_selection <- res$r2_income_ea3_wt/res$r2_income_ea3 - 1
prop_effect_selection <- scales::percent(prop_effect_selection, accuracy = 0.1)

prop_wt_fe <- res$r2_income_ea3_fe_wt/res$r2_income_ea3_fe - 1
prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)

res <- lapply(res, format, digits = 3)

From the meta-analysis of twin studies in @branigan2013variation, we take the
heritability of educational attainment to be around 0.4. This is the variance of
educational attainment explained by the "true" polygenic score for educational
attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
attainment from genetic data. We write the measured polygenic
score of educational attainment, "EA3" from @lee2018gene, as
\[
p=p^{*}+\eta.
\]

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}}=\frac{\sigma_{p}^{2}-\sigma_{\eta}^{2}}{\sigma_{p}^{2}}=\frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}\equiv\lambda,\label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.

We now consider the $R^2$ of $p$ on income, that is, the proportion
of variance in income that is explained by the true polygenic score
of educational attainment. Applying (\ref{eq:r2-ratio}) again gives
that
\begin{equation}
\frac{\textrm{plim}\hat{R}_{\textrm{income}}^{2}}{R_{\textrm{income}}^{2}}=\lambda.\label{eq:r2-ratio-income}
\end{equation}

We compute $\hat{R}_{\textrm{income}}^{2}$by regressing income group on EA3,
first weighting by age/qualification, and then by age/qualification weights
times respondent's number of children. The estimated $R^2$ are 
`r res$r2_income_ea3` and `r res$r2_income_ea3_wt` respectively. Applying
(\ref{eq:r2-ratio-income}) to both estimates gives the proportion of variance in
income explained by true polygenic score for education attainment (a) among
respondents as `r res$r2_income_true_psea`; (b) accounting for natural selection
as `r res$r2_income_true_psea_wt`. The second figure gives the counterfactual
estimate of $R_{\textrm{income}}^{2}$ after one generation of natural selection,
holding environmental conditions unchanged.

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ dues not differ between the two sets of
weighted regressions.

The estimate above uses a pooled regression of income on EA3, which may include
confounding environmental variation such as genetic nurture. Within-siblings
regression would give a smaller estimate of the effect of EA3. However, this
would also affect the reliability of the regression [@griliches1986errors],
making it hard to back out estimates of the true $R^2$. Nevertheless, we
can calculate the proportional increase in $R^2$ due to natural selection, again 
so long as the reliability coefficient is unaffected by
the weighting scheme. Running within-siblings regressions of EA3 on income group
gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this procedure estimates the proportional change in $R^2$ 
as `r prop_wt_fe`.

### Proof of \@ref(eq:r2-ratio)

Write 
\begin{align}
\hat{R}^{2} & =1-\frac{\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}}{\sum_{i=1}^{N}(EA_{i}-\bar{EA})^{2}}\label{eq:r-squared}
\end{align}

Using the fact that $\hat{\alpha}=\bar{EA}-\hat{\beta}\bar{p}$, where bars signify
sample averages, we can simplify the numerator in the above:

\[
\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\bar{EA}-\hat{\beta}\bar{p}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}
\]

Now, we can compute the following probability limit:

\begin{align}
 & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\nonumber \\
= & \text{plim}\frac{\sum_{i=1}^{N}\hat{\beta}^{2}(p_{i}-\bar{p})^{2}+(EA_{i}-\bar{EA})^{2}-2\hat{\beta}(p_{i}-\bar{p})(EA_{i}-\bar{EA})}{N}\nonumber \\
= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta\lambda\sigma_{p,EA},\label{eq:plim}
\end{align}

using the standard errors-in-variables formula $\textrm{plim}\hat{\beta}=\lambda\beta$,
and where $\sigma_{p,EA}$ is the covariance between $p$ and $EA$.

Write
\begin{align*}
\sigma_{p,EA} & =\textrm{Cov}(p^{*}+\eta,\alpha+\beta p^{*}+\varepsilon)\\
 & =0+\beta\sigma_{p^{*}}^{2}+\beta\textrm{Cov}(\eta,p^{*})+\textrm{Cov}(\eta,\varepsilon)\\
 & =\beta\sigma_{p^{*}}^{2}
\end{align*}

using the errors-in-variables assumption that $\eta$ is independent
of $p^{*}$ and $\varepsilon$. Plug this into (\ref{eq:plim}) to
give
\begin{align*}
 & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\\
= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\\
= & \beta^{2}\lambda\sigma_{p^{*}}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\textrm{, using \ensuremath{\sigma_{p^{*}}^{2}=\lambda\sigma_{p}^{2}},}\\
= & \sigma_{EA}^{2}-\beta^{2}\lambda\sigma_{p^{*}}^{2}
\end{align*}

Inserting this into (\ref{eq:r-squared}) gives
\begin{align*}
\textrm{plim}\hat{R}^{2} & =1-\frac{\sigma_{EA}^{2}}{\sigma_{EA}^{2}}+\frac{\beta^{2}\lambda\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
 & =\lambda\frac{\beta^{2}\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
 & =\lambda R^{2}
\end{align*}
since for bivariate regression $R^{2}$ is the square of the correlation
between $EA$ and $p^{*}$, which can be written $r=$$\beta\frac{\sigma_{p^{*}}^{2}}{\sigma_{EA}\sigma_{p^{*}}}$.
We thank an anonymous commenter on https://stats.stackexchange.com/questions/238878/how-do-errors-in-variables-affect-the-r2
for much of this proof.

@@

# Model proofs
